Relationships in the extremes and their influence on competition and coexistence: Supporting information

### 2022-01-10 paper/Paper.Rnw
edits leading up to editing Jasmin's new Methods content on the empirical system, also added new lines to that content to make it better tracked by git. Will edit that content next.

%***DAN: Dan made it to here on 2022 01 09 editing, continue tomorrow.

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence
Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
not been studied, the importance for coexistence of other aspects of climate change, including changes 
in means and measures of variability of environmental variables, has been studied in many systems, so we briefly 
review some of that large, related literature.
For instance, \cite{diPaola_2012} concluded that coexistence of drought sensitive,
flood resistant species with flood sensitive, drought resistant species in 
Mediteranean coastal swamp forests of central Italy would be jeopardized
by projected changes in mean rainfall for the region. In an interesting example
involving climate-change effects on herbivore-mediated competition in plants,
\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
higher altitudes of herbivore populations than of plant populations, newly arrived
herbivores at high altitudes can inflate species richness by reducing dominant plant species 
in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
all species they considered, but that changes in competition would more than offset these direct benefits,
and, as a result, some species would be extirpated. Numerous studies also take an ``event-focused'' 
perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
form of extreme weather events, can alter competitive interactions. For instance, 
\cite{White_2001} showed that single extreme heating events can alter competition 
between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
nutrient availability for plants, which is mediated by soil moisture content, can alter 
the potential for invasive species to enter an ecosystem and establish, a process that involves
the invader overcoming competition from resident species. 
Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
communities, via storage effects \citep{Adler_2006,  Angert_2009}.
Changes in variability and in extreme events are distinct from the changes in ATAs we 
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

@@

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs exist and are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATA plays a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, known mechanisms of coexistence. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical importance and its importance for real ecosystems.

@@

For a general model, the storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between invader and resident mean growth rates, when $i$ is the invader.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``invasion growth rate'' [IGR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on the invader or resident have been rendered independent.  
Following \cite{Ellner_2016,Ellner_2019}, a subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The notation $\Delta_i^{(EC)}$ for storage effects also follows \cite{Ellner_2019}.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the definition of $q_{ij}$. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions. Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.

@@

All computations were done in R on an ordinary laptop. Complete computer codes for the project are stored at
\url{https://github.com/jasminalbert/AsymmetricRelationshipsAndCoexistence.git}.

%Ideas while driving to pick up Max:
%1) In this paper with Jasmin you need to cite the two papers you just got from the professor at University of Massachusetts, Lowell. I mean the papers about changes in the special extent of heat waves: https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2020GL087097 and https://iopscience.iop.org/article/10.1088/1748-9326/ab4b41
%The second one is the one to cite
%2) Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model. Probably real environmental variables are actually related to vital rates in a non-linear and probably often threshold-like way. Threshold relationships can produce tail association, as we showed elsewhere. Changes in the means of true environmental variables may change where those distributions of environmental variables are compared to thresholds, which may change the nature of tail associations between the effects of those environmental variables.

@@

>>>>

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence
Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
not been studied, the importance for coexistence of changes 
in means and variability of environmental variables has been studied in many systems. We briefly 
review some of that literature.
\cite{diPaola_2012} concluded that coexistence of drought sensitive,
flood resistant species with flood sensitive, drought resistant species in 
Mediteranean coastal swamp forests of central Italy would be jeopardized
by projected changes in mean rainfall. 
\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
higher altitudes of herbivore populations than of plant populations, newly arrived
herbivores at high altitudes can inflate plant species richness by reducing dominant plant species 
in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
all species they considered, but that changes in competition would more than offset these direct benefits. 
Numerous studies take an ``event-focused'' 
perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
form of extreme weather events, can alter competitive interactions. For instance, 
\cite{White_2001} showed that single extreme heat events can alter competition 
between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
nutrient availability for plants, mediated by soil moisture content, can alter 
the potential for invasive species to enter an ecosystem and establish, a process that involves
the invader overcoming competition from resident species. 
Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
communities, via storage effects \citep{Adler_2006,  Angert_2009}.
Changes in variability and in extreme events are distinct from the changes in ATAs we 
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

@@

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATA plays a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence. This is to help 
illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using the lottery model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical importance and its importance for real ecosystems.

@@

For a general model, the storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between invader and resident mean growth rates, when $i$ is the invader.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``invasion growth rate'' [IGR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on the invader or resident have been rendered independent.  
Following \cite{Ellner_2016,Ellner_2019}, a subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The notation $\Delta_i^{(EC)}$ for storage effects also follows \cite{Ellner_2019}.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the definition of $q_{ij}$. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
Though some readers may find it slightly confusing to make simultaneous use of the variable 
$E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard and very common, so it would 
be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like
all operators, $\E(\cdot)$ is rendered upright.

@@

All computations were done in R on an ordinary laptop running Ubuntu Linux. 
Complete computer codes for the project are stored at
\url{https://github.com/jasminalbert/AsymmetricRelationshipsAndCoexistence.git}.

@@

%Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model. Probably real environmental variables are actually related to vital rates in a non-linear and probably often threshold-like way. Threshold relationships can produce tail association, as we showed elsewhere. Changes in the means of true environmental variables may change where those distributions of environmental variables are compared to thresholds, which may change the nature of tail associations between the effects of those environmental variables.

### 2022-01-10 paper/SupMat.Rnw
edits leading up to editing Jasmin's new Methods content on the empirical system, also added new lines to that content to make it better tracked by git. Will edit that content next.

For the lottery model, for $i \neq j$, we have
\begin{align}
\sC_j &= -\ln \left[ 1-\delta+\delta N \frac{\overline{B}_j}{B_1(t) N_1 (t) + B_2(t) N_2(t)}  \right]  \\
\sC_{j \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_j}{B_j(t)} \right] \\
\sC_{i \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_i}{B_j(t)} \right].
\end{align}
Both $\sC_{i \bs i}$ and $\sC_{j \bs i}$ are strictly monotonic functions of $B_j$, so the function
$B_j \mapsto \sC_{j \bs i}$ is invertible, and using its inverse provides an expression of $\sC_{i \bs i}$ 
in terms of $\sC_{j \bs i}$. One could write this expression explicitly and then differentiate it calculate $q_{ij}$,
but we instead compute $q_{ij}$ through implicit differentiation:
\begin{align}
\frac{\partial \sC_{i \bs i}}{\partial \sC_{j \bs i}} &= \left. \frac{\partial \sC_{i \bs i}}{\partial B_j} \middle/ \frac{\partial \sC_{j \bs i}}{\partial B_j} \right. \\
&= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{B_j}}{1-\delta+\delta \frac{\overline{B}_i}{B_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right).
\end{align}
Solving $\sC_{j \bs i}=0$ gives $B_j=\overline{B}_j$, so 
\begin{align}
q_{ij} &= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{\overline{B}_j}}{1-\delta+\delta \frac{\overline{B}_i}{\overline{B}_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right) \\
&= \frac{\overline{B}_i}{\overline{B}_j (1-\delta) +\overline{B}_i \delta},
\end{align}
as stated in the main text.

%***DAN: Change the below paragraph if needed. Some reasons it might be needed: Steve joins as a coauthor;
%Steve reads this and thinks it makes sense to change it because something is factually or presentationally 
%sub-optimal; or Dan or Jasmin actually develops a more rigorous understanding of scaling factors and based
%on that thinks something here is sub-optimal. Anyway, the "do both" approach seems like it can't go wrong,
%especially if Steve's prediction is correct that it won't make much difference.
%***DAN: You probably need to spend a bit more time with the cited papers to better understand this. 
We also use $q_{ij}=1$ as an alternative value, based on communication with Steve Ellner, for the following reasons.
First, as discussed in the supporting information of \cite{Ellner_2016}, and as further elaborated by 
\cite{Barabas_2018}, it takes very special circumstances for the scaling factors to exist and be unique, 
especially with size-structured models. Second, in empirically parameterized models, scaling factor values are 
often not robust to arbitrary modelling choices. 
Although the lottery model is not size structured and we do not empirically parameterize it, the above shortcomings of 
scaling factors, generally, suggest we should find alternatives to their use. The alternative 
suggested by Ellner for empirical applications is to do a series of pairwise invader-resident comparisons, each with an 
equal weighting of invader and resident $r$ values; for our context this amounts to setting scaling factors to $1$. 
We consider both approaches because the above-mentioned shortcomings do not apply to our case, but it seems 
possible that future applications will often set scaling factors to $1$, and we want our ideas to also be considered 
by researchers making that choice.

>>>>

For the lottery model, for $i \neq j$, we have
\begin{align}
\sC_j &= -\ln \left[ 1-\delta+\delta N \frac{\overline{B}_j}{B_1(t) N_1 (t) + B_2(t) N_2(t)}  \right]  \\
\sC_{j \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_j}{B_j(t)} \right] \\
\sC_{i \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_i}{B_j(t)} \right].
\end{align}
Both $\sC_{i \bs i}$ and $\sC_{j \bs i}$ are strictly monotonic functions of $B_j$, so the function
$B_j \mapsto \sC_{j \bs i}$ is invertible, and using its inverse provides an expression of $\sC_{i \bs i}$ 
in terms of $\sC_{j \bs i}$. One could write this expression explicitly and then differentiate it to calculate $q_{ij}$,
but we instead compute $q_{ij}$ through implicit differentiation:
\begin{align}
\frac{\partial \sC_{i \bs i}}{\partial \sC_{j \bs i}} &= \left. \frac{\partial \sC_{i \bs i}}{\partial B_j} \middle/ \frac{\partial \sC_{j \bs i}}{\partial B_j} \right. \\
&= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{B_j}}{1-\delta+\delta \frac{\overline{B}_i}{B_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right).
\end{align}
Solving $\sC_{j \bs i}=0$ gives $B_j=\overline{B}_j$, so 
\begin{align}
q_{ij} &= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{\overline{B}_j}}{1-\delta+\delta \frac{\overline{B}_i}{\overline{B}_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right) \\
&= \frac{\overline{B}_i}{\overline{B}_j (1-\delta) +\overline{B}_i \delta},
\end{align}
as stated in the main text.

%***DAN: Change the below paragraph if needed. Some reasons it might be needed: Steve joins as a coauthor - easy
%just means a change in wording at the top bit;
%Steve reads this and thinks it makes sense to change it because something is factually or presentationally 
%sub-optimal; or Dan or Jasmin actually develops a more rigorous understanding of scaling factors and based
%on that thinks something here is sub-optimal. Anyway, the "do both" approach seems like it can't go wrong,
%especially if Steve's prediction is correct that it won't make much difference.
%***DAN: Dan, you probably need to spend a bit more time with the cited papers to better understand this. 
We also use $q_{ij}=1$ as an alternative value, based on communication with Steve Ellner, for the following reasons.
First, as discussed in the supporting information of \cite{Ellner_2016}, and as further elaborated by 
\cite{Barabas_2018}, it takes very special circumstances for the scaling factors to exist and be unique, 
especially with size-structured models. Second, in empirically parameterized models, scaling factor values are 
often not robust to arbitrary modelling choices. 
Although the lottery model is not size structured and we do not empirically parameterize it, the above shortcomings of 
scaling factors, generally, suggest we should find alternatives to their use. The alternative 
suggested by Ellner for empirical applications is to do a series of pairwise invader-resident comparisons, each with an 
equal weighting of invader and resident $r$ values; for our context this amounts to setting scaling factors to $1$. 
We consider both approaches because the above-mentioned shortcomings do not apply to our case, but it seems 
possible that future applications will often set scaling factors to $1$, and we want our ideas to also be considered 
by researchers making that choice.
